{RandomRationalPoints} package for \emph{Macaulay2}

### 2020-12-21 FastLinAlg/Paper/RandomRationalPoints.tex
Examples Added, Paper Ready?

 There is an existing package called {\tt RandomPoints}, which we took inspiration from, which aims to find \emph{all} the rational points of a variety, our aim here is to find one point quickly, even if it is not rational.  
 We provide some functions that apply different strategies to generate random rational points for the given variety. We also provide functions that will expedite the process of determining properties of the singular locus of $X$ when combined with the functions provided in the package {\tt FastLinAlg}.

@@

	\begin{itemize}
		\item {\tt Replacement}: setting {\tt Replacement => Full} will mean that coordinates are replaced by a general degree $1$ form.  If {\tt Replacement => Binomial}, the coordinates are only changed to binomials, which can be much faster for certain applications. If {\tt Homogeneous => false}, then there will be constant terms, and we view $mx + b$ as a binomial.
		\item {\tt MaxCoordinatesToReplace}: the user can specify that only a specified number of coordinates should be non-monomial (assuming {\tt Homogeneous => true}).  
		This option is passed to {\tt randomCoordinateChange} by other functions that call it. 		
		\item {\tt Homogeneous}:  Setting {\tt Homogeneous => false} will cause degree zero terms to be added to modified coordinates (including monomial coordinates).		
	\end{itemize}

@@

\begin{example}	
	{{\small\color{blue}
	\begin{verbatim}

@@

\begin{example}
	{{\small\color{blue}
	\begin{verbatim}

@@

\begin{description}		
	\item[\tt Strategy => $\bullet$]   Here the $\bullet$ can be {\tt Default, BruteForce, LinearIntersection, GenericProjection} or {\tt HybridProjectionIntersection}.

@@

	Notice that speed, or success, varies depending on the strategy.

	\begin{example}\label{BruteForce}
		{{\small\color{blue}
		\begin{verbatim}

@@

	When calling the Strategy {\tt GenericProjection} or {\tt HybridProjectionIntersection} from {\tt randomPoints}, this option denotes the number of trials before giving up.  This option is also passed to randomPoints by other functions.

@@

	\begin{example}
	We re-compute \cref{BruteForce} this time specifying more attempts.
		{{\small\color{blue}
		\begin{verbatim}
	i1: R = ZZ/101[x,y,z];

@@

	When working over very small fields, frequently {\tt BruteForce} is most efficient.  This is not surprising as there may not be many points to check, and a surprisingly large percentage of points can be rational in hypersurfaces and complete intersections.  However, if the field size is larger, {\tt BruteForce} will perform poorly.  Other strategies work differently on different examples, and even the same strategy can sometimes work very quickly even if it typically works very slowly.
	\begin{example}
		{{\small\color{blue}
		\begin{verbatim}

@@

	-- the indices of the linearly independent rows of the matrix extracted from $M$ using (2),

	-- a random $n\times n$ submatrix of $M$ that has full rank at $P$.

@@

	\item[\tt Verbose => Boolean]
		 Set the option {\tt Verbose => true} to turn on verbose output.  This may be useful in debugging or in determining why an computation is running slowly.

@@

@@

\vspace{1em}
\noindent where $n,M,I$ are same as before. This function finds a submatrix of size $n\times n$ using {\tt findANonZeroMinor};  
it extracts the last entry of the output, finds its determinant and
adds it to the ideal $I$, thus extending $I$.

%\vspace{1em}
As again, we can have the following additional options:
\begin{description}
\item[\tt Strategy => Symbol] This specifies which strategy to use when calling {\tt randomPoints}.
\item[\tt Homogeneous => Boolean] This is passed in calls to {\tt randomPoints}.
\item[\tt Verbose => Boolean] This turns on or off verbose output (see \Cref{findANonZeroMinor}).
\item[\tt MinorPointAttempts => ZZ] This controls how many points at which to check the rank of the matrix.
\end{description}

One use for this function can be in showing that certain rings are $(R_1)$ (regular in codimension $1$).
Consider the following example which is $(R_1)$ where computing the dimension of the singular locus takes around $30$ seconds as there are $15500$ minors of size $4 \times 4$ in the associated $7 \times 12$ Jacobian matrix.  However, we can use this function to quickly find interesting minors.
\begin{example}

>>>>

 There is an existing package called {\tt RandomPoints}, which we took inspiration from, which aims to find \emph{all} the rational points of a variety; our aim here is to find one point quickly, even if it is not rational.  
 We provide some functions that apply different strategies to generate random rational points for the given variety. We also provide functions that will expedite the process of determining properties of the singular locus of $X$ when combined with the functions provided in the package {\tt FastLinAlg}.

@@

	\begin{itemize}
		\item {\tt Replacement}: Setting {\tt Replacement => Full} will mean that coordinates are replaced by a general degree $1$ form.  If {\tt Replacement => Binomial}, the coordinates are only changed to binomials, which can be much faster for certain applications. If {\tt Homogeneous => false}, then there will be constant terms, and we view $mx + b$ as a binomial.
		\item {\tt MaxCoordinatesToReplace}: the user can specify that only a specified number of coordinates should be non-monomial (assuming {\tt Homogeneous => true}).  
		This option is passed to {\tt randomCoordinateChange} by other functions that call it. 		
		\item {\tt Homogeneous}:  Setting {\tt Homogeneous => false} will cause degree zero terms to be added to modified coordinates (including monomial coordinates).		
	\end{itemize}

@@

~~

\begin{example}	
~~

{{\small\color{blue}
	\begin{verbatim}

@@

~~

\begin{example}~~

	{{\small\color{blue}
	\begin{verbatim}

@@

\begin{description}		
	\item[\tt Strategy => $\bullet$]   Here the $\bullet$ can be {\tt Default, BruteForce, LinearIntersection, \\ GenericProjection} or {\tt HybridProjectionIntersection}.

@@

	Notice that the speed, or success, varies depending on the strategy.

	\begin{example}\label{BruteForce}
		~~

		{{\small\color{blue}
		\begin{verbatim}

@@

	When calling the Strategy {\tt GenericProjection} or \\
	{\tt HybridProjectionIntersection} from {\tt randomPoints}, this option denotes the number of trials before giving up.  This option is also passed to {\tt randomPoints} by other functions.

@@

	We re-compute \cref{BruteForce} this time specifying more attempts.

		{{\small\color{blue}
		\begin{verbatim}
	i1: R = ZZ/101[x,y,z];

@@

	When working over very small fields, frequently {\tt BruteForce} is most efficient.  This is not surprising as there may not be many points to check, and a surprisingly large percentage of points can be rational in hypersurfaces and complete intersections.  However, if the field size is larger, {\tt BruteForce} will perform poorly.  Other strategies work differently on different examples, and even the same strategy can sometimes work very quickly even if it typically works very slowly.

	~

	\begin{example}
		~~

		{{\small\color{blue}
		\begin{verbatim}

@@

	-- the indices of the linearly independent rows of the matrix extracted from $M$ in the above step,

	-- a random $n\times n$ sub-matrix of $M$ that has full rank at $P$.

@@

	\item[\tt Verbose => Boolean]
		 Set the option {\tt Verbose => true} to turn on verbose output.  This may be useful in debugging or in determining why a computation is running slowly.

@@

\vspace{1em}

\begin{example}~~

	~~
	{{\small\color{blue}
		\begin{verbatim}
		i3 : R = ZZ/5[x,y,z]

		o3 = R

		o3 : PolynomialRing

		i4 : I = ideal(random(3,R)-2, random(2,R))

		              3     2        2    3    2              2        2       2
		o4 = ideal (2x  - 2x y + 2x*y  + y  + x z - 2x*y*z + y z - 2x*z  + 2y*z 
		-------------------------------------------------------------------
		   3                      2
		- z  - 2, - 2x*y - x*z - z )

		o4 : Ideal of R

		i5 :  M = jacobian(I)

		o5 = {1} | x2+xy+2y2+2xz-2yz-2z2   -2y-z |
		     {1} | -2x2-xy-2y2-2xz+2yz+2z2 -2x   |
		     {1} | x2-2xy+y2+xz-yz+2z2     -x-2z |

		             3       2
		o5 : Matrix R  <--- R

		i6 : findANonZeroMinor(2,M,I, Strategy => GenericProjection)

		o6 = ({-2, 1, 1}, {0, 1}, {0, 1}, {1} | x2+xy+2y2+2xz-2yz-2z2   -2y-z |)
		{1} | -2x2-xy-2y2-2xz+2yz+2z2 -2x   |

@@

\vspace{1em}
\noindent where $n,M,I$ are same as before. This function finds a submatrix of size $n\times n$ using \\{\tt findANonZeroMinor};  
it extracts the last entry of the output, finds its determinant and
adds it to the ideal $I$, thus extending $I$.

%\vspace{1em}
As before, we can have the following additional options:
\begin{description}
\item[\tt Strategy => Symbol] This specifies which strategy to use when calling {\tt randomPoints}.
\item[\tt Homogeneous => Boolean] This is passed in calls to {\tt randomPoints}.
\item[\tt Verbose => Boolean] This turns on or off verbose output (see \Cref{findANonZeroMinor}).
\item[\tt MinorPointAttempts => ZZ] This controls how many points at which to check the rank of the matrix.
\end{description}

\vspace{1em}
\begin{example}~~

	~~

	{{\small\color{blue}
\begin{verbatim}
i2 : R = ZZ/5[x,y,z]

i3 : I = ideal(random(3,R)-2, random(2,R))

                3    2       2     3     2              2       2     3
o3 = ideal (- 2x  + x y - x*y  - 2y  - 2x z + 2x*y*z + y z + y*z  - 2z 
-------------------------------------------------------------------
         2            2                 2
- 2, - 2x  - 2x*y + 2y  + x*z - 2y*z - z )

i4 : M = jacobian(I)

o4 = {1} | -x2+2xy-y2+xz+2yz    x-2y+z   |
     {1} | x2-2xy-y2+2xz+2yz+z2 -2x-y-2z |
     {1} | -2x2+2xy+y2+2yz-z2   x-2y-2z  |

             3       2
o4 : Matrix R  <--- R

i5 : extendIdealByNonZeroMinor(2,M,I, Strategy => LinearIntersection)

                3    2       2     3     2              2       2     3
o5 = ideal (- 2x  + x y - x*y  - 2y  - 2x z + 2x*y*z + y z + y*z  - 2z 
-------------------------------------------------------------------
         2            2                 2   3    2        2    3  
- 2, - 2x  - 2x*y + 2y  + x*z - 2y*z - z , x  + x y + 2x*y  - y  +
-------------------------------------------------------------------
  2       2    3
2x z + y*z  - z )

o5 : Ideal of R
\end{verbatim}	
}}	
\end{example}

One use for this function can be in showing that certain rings are $(R_1)$ (regular in codimension $1$).
Consider the following example which is $(R_1)$ where computing the dimension of the singular locus takes around $30$ seconds as there are $15500$ minors of size $4 \times 4$ in the associated $7 \times 12$ Jacobian matrix.  However, we can use this function to quickly find interesting minors.

\vspace{1em}

\begin{example}~~

	~~
	{{\small\color{blue}
	\begin{verbatim}
		i2 : T = ZZ/101[x1,x2,x3,x4,x5,x6,x7];

		i3 : I =  ideal(x5*x6-x4*x7,x1*x6-x2*x7,x5^2-x1*x7,x4*x5-x2*x7,x4^2-x2*x6,x1*x4-x2*x5,
		x2*x3^3*x5+3*x2*x3^2*x7+8*x2^2*x5+3*x3*x4*x7-8*x4*x7+x6*x7,
		x1*x3^3*x5+3*x1*x3^2*x7+8*x1*x2*x5+3*x3*x5*x7-8*x5*x7+x7^2,
		x2*x3^3*x4+3*x2*x3^2*x6+8*x2^2*x4+3*x3*x4*x6-8*x4*x6+x6^2,
		x2^2*x3^3+3*x2*x3^2*x4+8*x2^3+3*x2*x3*x6-8*x2*x6+x4*x6,
		x1*x2*x3^3+3*x2*x3^2*x5+8*x1*x2^2+3*x2*x3*x7-8*x2*x7+x4*x7,
		x1^2*x3^3+3*x1*x3^2*x5+8*x1^2*x2+3*x1*x3*x7-8*x1*x7+x5*x7);

		o3 : Ideal of T

		i4 : M = jacobian I;

		             7       12
		o4 : Matrix T  <--- T

		i5 : i = 0;

		i6 : J = I;

		o6 : Ideal of T

		i7 : elapsedTime(while (i < 10) and dim J > 1 do (
                                i = i+1; 
                                J = extendIdealByNonZeroMinor(4, M, J)));
		-- 0.903328 seconds elapsed

		i8 : dim J

		o8 = 1

		i9 : i

		o9 = 5

	\end{verbatim}
}}
